Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
update intro

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and an genetic components [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to determine the factors that cause mismatches between the geographical and the ecological rear-edge [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].

@@

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be severely threatened by climate change. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct the forest history of our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local‐scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

>>>>

The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been often assumed that geographically marginal populations represent ecological marginal populations. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].

Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009]. Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010]. Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge. The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005; @VilaCabrera2019].

It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017]. In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects]. The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse]. For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests]. Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica]. In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range. Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].

@@

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjectionsa; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
